Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
Beginning pass through spitting section.

Although the type system of \lang\ ensures that we can perform all first-stage
computations before any second-stage ones, actually evaluating terms in this
fashion is rather involved, because stages are syntactically interleaved
within terms. In this section, we present a \emph{splitting algorithm} which
statically converts a term in \lang\ into a pair of 
syntactically % I don't like this word but temporarily here for continuity
separate programs corresponding to the two stages of computation in that term.

@@

More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits into a syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$), which
can also be reduced to $q$ (via the new relation $\mathcal{P} \sepredtwosym q$).
We enforce the ``syntactically separated'' condition by saying that $\mathcal{P}$ 
can only have one form, namely $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term encoding all the
first-stage computation that was in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term encoding all the
second-stage computation that was in $e$.  
Likewise, $\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and plugs it in for $l$:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
Here we call $b$ the {\em boundary value}, as it represents the communication at the boundary between stages.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}),
we must also provide a way to split those as well.
This is given by the new relation $e \splitonesym \mathcal{P}$,
where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computation that was in $e$, 
and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the
second-stage computation that was in $e$.

Similar to above, we also define a $\sepredtwosym$ relation 
which can be used to evaluate these seperated world \bbonep terms.
One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, consider the term
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
which reduces (via $\redonesym$) to the residual table and partial value
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}
Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.

We resolve this mismatch by introducing a translation, called {\em masking} and written $\rtab \xi v \vsplito \mathcal{V}$,
which converts residual tables and partial values---like the those above---into a syntactically separated version $\mathcal{V}$.
As before, we enforce the ``syntactically separated'' property structurally,
saying that $\mathcal{V}$ must have the form $\mval i q$ (called a {\em masked value}) where 
$i$ is a monostage value representing the first-stage part of $v$,
and $q$ is a monostage term representing all of the second-stage computation in $\xi$ and $v$.
Note that masking is a technical device for the purpose of stating correctness of splitting at \bbonem; 
it is not part of the splitting algorithm itself.

With this machinery in place, we can define $\sepredonesym$ as just creating masked values with the single rule:
\[
\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
Here we see the primary difference between world \bbtwo\ terms and world \bbonem\ terms.
Since $\coltwo e A$ reduces to an entirely second-stage result, 
its first-stage subcomputations only exist to internally pass their value to the
second stage;
but since $\colmix e A$ reduces to a multistage result, 
its first-stage subcomputations exist {\em both} to pass their value to the
second stage {\em and} to be present in the result.
Correspondingly, in the $\sepredtwosym$ rule, the precomptuation $p$ reduces only to a boundary value which is passed to the resumer;
but in the $\sepredonesym$ rule, the combined term $c$ reduces to a tuple containing both the immediate result and the boundary value,
with only the latter being passed to the resumer.

@@

This section proceeds by defining masking ($\vsplito$), 
then using masking to motivate the definition of \bbonem-translation ($\splitonesym$),
and finally coming back to define \bbtwo-translation ($\splittwosym$).

>>>>

%Although the type system of \lang\ ensures that we can perform all first-stage
%computations before any second-stage ones, actually evaluating terms in this
%fashion is rather involved, because stages are syntactically interleaved
%within terms.

In this section, we present a \emph{splitting algorithm} which
statically converts a term in \lang\ into a pair of 
%syntactically % I don't like this word but temporarily here for continuity
separate, monostaged programs that correspond to the first and second stage computations in that term.

@@

More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits into a syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$) which
can also be reduced to $q$ (via the new relation $\mathcal{P} \sepredtwosym q$).
By ``syntactically separated'' we mean that $\mathcal{P}$ 
has the form $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term representing all the
first-stage computations in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term representing all the
second-stage computations in $e$.  
$\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and substitutes the resulting value in for $l$ in the resumer:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}), splitting is also defined on 
\bbonem\ terms. %we must also provide a way to split those as well.
This is given by the new relation $e \splitonesym \mathcal{P}$,
where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computations in $e$, 
and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the
second-stage computations in $e$.

Similar to above, we also define a $\sepredtwosym$ relation 
which can be used to evaluate these seperated world \bbonep\ terms.
One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, consider the term
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
which reduces (via $\redonesym$) to the residual table and partial value
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}
Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.

We resolve this mismatch by introducing a translation, called {\em masking} and written $\rtab \xi v \vsplito \mathcal{V}$,
which converts residual tables and partial values---like those in the example above---into a syntactically separated version $\mathcal{V}$.
As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value}) where 
$i$ is a monostage value representing the first stage part of $v$,
and $q$ is a monostage term representing all second-stage computations in $\xi$ and $v$.
Note that masking is a technical device used to state the correctness of splitting at \bbonem; 
it is not part of the splitting algorithm itself.

Given this definition of masking, we define $\sepredonesym$ by a single rule that creates masked values:
\[
\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
Here we see the primary difference between world \bbtwo\ terms and world \bbonem\ terms.
Since $\coltwo e A$ reduces to an entirely second-stage result, 
its first-stage subcomputations only exist to internally pass their value to the
second stage;
but since $\colmix e A$ reduces to a multistage result, 
its first-stage subcomputations exist {\em both} to pass their value to the
second stage {\em and} to be present in the result.
Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$ reduces only to a boundary value which is passed to the resumer;
but in the $\sepredonesym$ rule, the combined term $c$ reduces to a tuple containing both the immediate result and the boundary value,
with only the latter being passed to the resumer.

@@

This section proceeds by defining masking ($\vsplito$), 
then using masking to motivate the definition of \bbonem-term splitting ($\splitonesym$),
and finally coming back to define \bbtwo-term splitting ($\splittwosym$).
